Splitting it up: the \pkg{spduration} split-population duration regression package

### 2015-12-18 manuscript.Rmd
fix typos and footnote 3

---
author:
  - name: Andreas Beger
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: ab428@duke.edu
    url: something
  - name: Daniel W. Hill, Jr.
    affiliation: University of Georgia
    address: >
      First line
      Athens, GA, USA
    email: dwhill@uga.edu
  - name: Nils W. Metternich
    affiliation: University College London
    address: >
      First line
      London, UK
    email: n.metternich@ucl.ac.uk
  - name: Michael D. Ward
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: mw160@duke.edu
title:
  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration models. The package includes estimation and several post-estimation methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [regression, rare events, "\\proglang{R}"]
  plain:     [regression, rare events, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

@@

However, there are conditions under which not all individuals have the same underlying risk to experience a specific outcome and might not even be at risk at all. Consider the risk of acquiring a viral infection like the common flu. Let us initially assume that everyone is at risk of infection, and individual behavior (e.g. good hygiene) and structural factors (e.g. workplace) determine whether individuals catch the flu. Under these assumptions a standard duration model and should provide us with efficient estimates of how different behavioral and structural factors impact on the general baseline risk. However, in the overall population there might be individuals that are immune to the flu, because they received vaccination, had the specific flu virus in the past, or have some other characteristic that makes it impossible for them to attract the decease. In this instance we have two underlying populations: An at risk population and an immune one. If the immune population is relatively large, estimates using a standard duration model will be biased and predictions inaccurate.

@@

Duration models, where baseline risk is modeled by some distribution of time, were originally developed in health and demographic research and grew naturally from life tables and survival records for medical patients. Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-parametric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest. This assumption may be violated in many substantive areas and empirical applications, where a sub-population of units or individuals will never experience an event, and thus are effectively ``cured''.

The insight that populations might be split in regard to their baseline risk has been formulated as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients. In their application some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds. The development of duration methods in health and medicine has also shaped the terminology conventionally used for such model. For example, split-population duration models are also referred to as cure rate models. Similarly, the basic concepts like survival and failure rates reference to the survival of humans.

@@

In order to illustrate the package functionality, we examine a model of coups d`etat in @belkin2003toward. Belkin and Schofer's paper lends itself to re-analysis with a split-populuation duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in an at-risk regime. They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors. Using language which fits nicely with the class of models described above, they write "[t]riggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup. Hence, triggers should not be equated with coup risk. Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk" [@belkin2003toward, p. 598].

For their empirical test Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups. However, they implement a logistic regression model, which does not capture the process as described in their quote above. This is because the logistic regression model assumes that all observations are at risk for a coup (i.e., the probability of a coup is non-zero for all observations). Since their structural coup risk indicator is developed precisely to distinguish between at risk and ``immune'' cases, the split-population model allows one to examine whether the indicator effectively does so.

@@

For the test set it is recommended to add the duration variables and *then* subset the test set. Since the test set is later in time than the training set we do not have to worry about contamination of the risk coding, but if we subset the data before building the duration variables we will start all duration counters at 1996, when in fact we can safely use the previous historic coup information. To do this we need to build the duration variables first:

@@

Furthermore, in out-of-sample testing based on some kind of data partitioning scheme, this coding scheme can lead to unintentioned contamination of in-sample cases with knowledge of out-of-sample failures through the risk coding for failed spells. This leads to two recommendations. First, the duration data (`add_duration`) should not be built until after partitioning the data. Secondly, and as a result of the first point, data should be partitioned in a scheme that samples either by spell or block-wise, e.g. by witholding the last **x** years of data, in order to properly build the duration variables.

@@

Many outcomes of interest are rare events. For example, coups, war onset, or mass killings are exceedingly rare events when considering that most countries do not experience high levels of violence. Figure \ref{rates} shows a few examples of published research that models binary outcomes. @fearon2003ethnicity is a widely cited study of civil war onset that uses yearly observations of all major countries; @beger2014ensemble is an example of the positive rates when moving to monthly data for a similar set of countries. The positive rates range from 5 to less than 0.2\% of all data points.

>>>>

---
author:
  - name: Andreas Beger
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: ab428@duke.edu
    url: something
  - name: Daniel W. Hill, Jr.
    affiliation: University of Georgia
    address: >
      Department of International Affairs, University of Georgia
      Athens, GA, USA
    email: dwhill@uga.edu
  - name: Nils W. Metternich
    affiliation: University College London
    address: >
      Department of Political Science, University College London
      London, UK
    email: n.metternich@ucl.ac.uk
  - name: Shahryar Minhas
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: sfm12@duke.edu
  - name: Michael D. Ward
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: mw160@duke.edu
title:
  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration models. The package includes estimation and several post-estimation methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [regression, rare events, "\\proglang{R}"]
  plain:     [regression, rare events, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

@@

However, there are conditions under which not all individuals have the same underlying risk to experience a specific outcome and might not even be at risk at all. Consider the risk of acquiring a viral infection like the common flu. Let us initially assume that everyone is at risk of infection, and individual behavior (e.g. good hygiene) and structural factors (e.g. workplace) determine whether individuals catch the flu. Under these assumptions a standard duration model should provide us with efficient estimates of how different behavioral and structural factors impact on the general baseline risk. But, there might be individuals that are immune to the flu in the overall population--because they received vaccination, had the specific flu virus in the past, or have some other characteristic that makes it impossible for them to attract the disease. In this instance we have two underlying populations: an at risk population and an immune one. If the immune population is relatively large, estimates using a standard duration model will be biased and predictions inaccurate.

@@

Regular duration models, where baseline risk is modeled by some distribution of time, were originally developed in health and demographic research and grew naturally from life tables and survival records for medical patients. Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-parametric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest. This assumption may be violated in many substantive areas and empirical applications, where a sub-population of units or individuals will never experience an event, and thus are effectively ``cured''.

The insight that populations might be split in regard to their baseline risk has been formulated as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients. In their application some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds. The development of duration methods in health and medicine has also shaped the terminology conventionally used for such models. For example, split-population duration models are also referred to as cure rate models. Similarly, the basic concepts like survival and failure rates reference the survival of humans.

@@

In order to illustrate the package functionality, we examine a model of coups d`etat in @belkin2003toward. Belkin and Schofer's paper lends itself to re-analysis with a split-population duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in an at-risk regime. They argue that many countries never experience coups because coups are effectively impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors. Using language which fits nicely with the class of models described above, they write "[t]riggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup. Hence, triggers should not be equated with coup risk. Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk" [@belkin2003toward, p. 598].

For their empirical test Belkin and Schofer develop a measure of "structural coup risk" which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups. However, they implement a logistic regression model, which does not capture the process as described in their quote above. This is because the logistic regression model assumes that all observations are at risk for a coup (i.e., the probability of a coup is non-zero for all observations). Since their structural coup risk indicator is developed precisely to distinguish between at risk and "immune" cases, the split-population model allows one to examine whether the indicator effectively does so.

@@

For the test set it is recommended to add the duration variables and \textit{then} subset the test set. Since the test set is later in time than the training set we do not have to worry about contamination of the risk coding, but if we subset the data before building the duration variables we will start all duration counters at 1996, when in fact we can safely use the previous historic coup information. To do this we need to build the duration variables first:

@@

Furthermore, as the example above shows, in out-of-sample testing based on some kind of data partitioning scheme, this coding scheme can lead to unintentional contamination of in-sample cases with knowledge of out-of-sample failures through the risk coding for failed spells. This leads to two recommendations. First, data should be partitioned spell or block-wise, e.g. by withholding the last *x* years of data, and not randomly. Second, given the two concerns of left-censoring and non-independence induced through the duration and risk coding, care should be taken to ensure that duration data are built without access to future information in another data partition.

@@

Many outcomes of interest are rare events. For example, coups, war onset, or mass killings are exceedingly rare events when considering that most countries do not experience high levels of violence. Figure \ref{rates} shows a few examples of published research that models binary outcomes. @fearon2003ethnicity is a widely cited study of civil war onset that uses yearly observations of all major countries; @beger2014ensemble is an example of the positive rates when moving to monthly data for a similar set of countries. The positive rates range from 5 to less than 0.2\% of all data points.[^3]

[^3]: For a discussion of the difficulties rare events can pose for prediction see @king2001logistic and @king2001explaining.
